---
title: "First Presentation"
author: "Rohit Padebettu"
date: "10/15/2016"
output: ioslides_presentation
---

## R Markdown

This is an R Markdown presentation. Markdown is a simple formatting syntax for authoring HTML, PDF, and MS Word documents. For more details on using R Markdown see <http://rmarkdown.rstudio.com>.

When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document.

## Slide with Bullets

- Bullet 1
- Bullet 2
- Bullet 3

## Slide with R Code and Output

```{r}
summary(cars)
```

## My slide xtable

`head(mtcars)`

```{r ,echo=FALSE, results='asis'}
library(xtable)
head<-xtable(head(mtcars),digits = 2, align = c(rep("l", 4), rep("c", 4), rep("r", 4)))
print(head,"html")
```

## My slide kable

```{r kable ,echo=FALSE, results='asis'}
suppressWarnings(library(knitr))
head<-kable(head(mtcars),digits = 2)
print(head)
```

## Slide with Plot

```{r, echo=FALSE}
plot(cars)
```

