---
title: "Course Project - Storm Data"
author: "Rohit Padebettu"
date: "10/17/2016"
output: html_document
---

## Loading libraries
```{r,echo= TRUE,warning=FALSE}
suppressPackageStartupMessages(library(R.utils))         # for unzipping the file
suppressPackageStartupMessages(library(data.table))      # for quick reading and subsetting
suppressPackageStartupMessages(library(lubridate))       # for date transformations
suppressPackageStartupMessages(library(dplyr) )          # for data transformations
suppressPackageStartupMessages(library(ggplot2))         # for plots
suppressPackageStartupMessages(library(ggthemes))        # themes for plots
suppressPackageStartupMessages(library(plotly) )         # for fancier interactive plots
```

```{r DownloadData,eval=FALSE}
## Downloading file and decompressing
setwd("./Final Course Project/")
url<-c("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2")

## Run the following code once
download.file(url,"./Data/StormData.csv.bz2",method = "curl")
bunzip2("./Data/StormData.csv.bz2",remove = FALSE)
```

## Reading into a data.table
```{r ReadData,cache=TRUE,warning=FALSE}
StormData<-fread("./Data/StormData.csv",header = TRUE,strip.white = TRUE)
```

### Converting to Date format to allow subsetting and Getting only post 1992 data
```{r DataSub}
StormData$BgnDate <- as_date(mdy_hms(StormData$BGN_DATE))
StormData$EndDate <- as_date(mdy_hms(StormData$END_DATE))

StormData_sub <- StormData[year(BgnDate)>1992,
                           .(BgnDate,EndDate,REFNUM,
                            STATE,STATE__,COUNTY,COUNTYNAME,
                            EVTYPE,FATALITIES,INJURIES,
                            PROPDMG,PROPDMGEXP,CROPDMG,CROPDMGEXP)]

Data_df<-as.data.frame(StormData_sub)
```

## Computing the Total Dollar Damage per event

### Function used to decode the EXP in the dataset
```{r fDecode}
decoder<-function(v){
    
    if(length(v[2])==0){return (1)}
    else if(v[2]%in% c("H","h")) {return(100)}
    else if(v[2]%in% c("K","k")) {return(1000)}
    else if(v[2]%in% c("M","m")) {return(1000000)}
    else if(v[2]%in% c("B","b")) {return(1000000000)}
    else if(v[2]%in% c("0","1","2","3","4","5","6","7","8","9")) {10^as.numeric(v[2])}
    else if(v[2]=="") {return(1)} 
    else return(1)
}
```

```{r Prepare,cache=TRUE}
### Preparing the dataset for analysis
PROPEXP<-apply(Data_df[,c(11,12)],1,decoder)
CROPEXP<-apply(Data_df[,c(13,14)],1,decoder)
Data_df<-cbind(Data_df,PROPEXP,CROPEXP)
Data_df <- mutate(Data_df,Tot_damage = (PROPDMG*PROPEXP)+(CROPDMG*CROPEXP))

```

## Top Event Types which cause the most economic damage

### Select the top 50 percentile of event data
```{r EcoDmg}
n=50
Data_calc<-Data_df%>%
    filter(Tot_damage > quantile(Tot_damage, prob = 1 - n/100))%>%
    select(Date=BgnDate,Event=EVTYPE ,Damage = Tot_damage)%>%
    mutate(Event = as.factor(toupper(trim(Event))))%>%
    group_by(Event)%>%
    summarize(EventDamage = sum(Damage)/1000000000)%>%
    arrange(desc(EventDamage))
```

```{r EcoDmgPlot,fig.height=5,fig.width=8}
### Select top 10 Event Types
Data_top10<-Data_calc[c(1:10),]

### Plot the results
g1<-ggplot(Data_top10,aes(x=factor(Event),y=EventDamage,fill=Event))+
    geom_bar(stat="identity",col="black",width=0.5)+
    coord_flip()+
    guides(fill=FALSE)+
    ylab("Economic Damage(in $ billions)")+
    xlab("Top 10 Event Types")+
    ggtitle("Most Economically Damaging Event Types")+
    theme_gdocs()
print(g1)
```


## Yearly Economic Damage
```{r YearEcoDmg}
Data_year<-Data_df%>%
    select(Date=BgnDate,Event=EVTYPE,Damage = Tot_damage)%>%
    group_by(Year = year(Date))%>%
    summarize(YearlyDamage = sum(Damage)/1000000000)%>%
    arrange(Year)
```

```{r YearEcoDmgPlot,fig.height=5,fig.width=8}
### Plot the results
g2<-ggplot(Data_year,aes(x=factor(Year),y=YearlyDamage,fill=Year))+
    geom_bar(stat="identity",col="black",width=0.5)+
    coord_flip()+
    guides(fill=FALSE)+
    ylab("Economic Damage(in $ billions)")+
    xlab("Years")+
    ggtitle("Yearly Economic Damage")+
    theme_gdocs()
ggplotly(g2)
```

## Top Event Types which cause the most human damage
```{r HumanDmg}
Data_hum<-Data_df%>%
    select(Date=BgnDate,Event=EVTYPE,FATALITIES,INJURIES)%>%
    mutate(Event = as.factor(toupper(trim(Event))))%>%
    group_by(Event)%>%
    summarize(HumanDamage=sum(FATALITIES,INJURIES),
              Deaths=sum(FATALITIES))%>%
    mutate(Intensity = Deaths/HumanDamage)%>%
    arrange(desc(HumanDamage))
```

### Select top 10 Event Types

```{r HumanDmgPlot,fig.height=5,fig.width=8}
Data_top10_h<-Data_hum[c(1:10),]

### Plot the results

g3<-ggplot(Data_top10_h,aes(x=factor(Event),y=HumanDamage,fill=Intensity))+
    geom_bar(stat="identity",col="white",width=0.5)+
    coord_flip()+
    ylab("Human Damage(Fatalities+Injuries)")+
    xlab("Top 10 Event Types")+
    ggtitle("Event types causing most human damage")+
    theme_fivethirtyeight()
print(g3)
```
